OptiFit: an improved method for fitting amplicon sequences to existing OTUs

### 2022-01-13 paper/paper.Rmd

First, we tested whether OptiFit performed as well as _de novo_ clustering when
using the split dataset strategy with half of the sequences selected for
the reference by a simple random sample (a 50% split)
(Figure \ref{fig:results_sum}; self-split).
OTU quality was similar to that from OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
In closed reference mode, OptiFit was able to cluster `r frac_fit_split`% of query
sequences to reference OTUs with the split strategy, a great improvement over
the average `r frac_fit_gg`% of sequences clustered to the Greengenes database.
In terms of run time, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and
`r sec_open_fit_split_vs_clust`%, respectively.
The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster
than the database strategy in closed reference mode and
`r sec_open_fit_split_vs_db`% faster in open reference mode.
Thus, reference clustering with the split dataset strategy creates as high
quality OTUs as _de novo_ clustering yet at a faster run time, and fits far more
query sequences than the database strategy.

@@

```{r assert_split_fracs, include = FALSE}
close_enough(cv_fit_split_mcc_human_simple, 0, tol = 10^-3)
assert_that(sec_fit_split_human_simple_1 > sec_fit_split_human_simple_9)
assert_that(frac_fit_split_human_simple_1 < frac_fit_split_human_simple_9)
```

While we initially tested this strategy using a 50% split of the data into
reference and query fractions, we next investigated whether there was
an optimal reference fraction size.
To identify the best reference size, reference sets with 10% to 90% of the
sequences were created, with the remaining sequences used for the query
(Figure \ref{fig:results_split}).
OTU quality was remarkably consistent across reference fraction sizes.
For example, splitting the human dataset 100 times yielded a coefficient of
variation (i.e. the standard deviation divided by the mean) of
`r cv_fit_split_mcc_human_simple` for the MCC score across all fractions.
Run time generally decreased as the reference fraction increased; for the human
dataset, the median run time was `r sec_fit_split_human_simple_1` seconds with
10% of sequences in the reference and `r sec_fit_split_human_simple_9` seconds
with 90% of sequences in the reference.
In closed reference mode, the fraction of sequences that mapped increased as the
reference size increased; for the human dataset, the median fraction mapped was
`r frac_fit_split_human_simple_1` with 10% of sequences in the reference and
`r frac_fit_split_human_simple_9` with 90% of sequences in the reference.
These trends held for the other datasets as well.
Thus, the reference fraction did not affect OTU quality in terms of MCC score,
but did affect the run time and the fraction of sequences that mapped during the
closed reference clustering.

>>>>

First, we tested whether OptiFit performed as well as _de novo_ clustering when
using the split dataset strategy with half of the sequences selected for
the reference by a simple random sample (a 50% split)
(Figure \ref{fig:results_sum}; self-split).
OTU quality was similar to that from OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
In closed reference mode, OptiFit was able to cluster `r frac_fit_split`% of query
sequences to reference OTUs with the split strategy, a great improvement over
the average `r frac_fit_gg`% of sequences clustered to the Greengenes database.
In terms of run time, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and
`r sec_open_fit_split_vs_clust`%, respectively.
Random access memory (RAM) usage was similar, with OptiFit requiring slightly 
more RAM in gigabytes than OptiFit. 
Open and closed reference OptiFit required `r mem_open_fit_split_vs_clust`% and
`r mem_closed_fit_split_vs_clust`% more RAM than OptiClust, respectively 
(data not shown).
The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster
than the database strategy in closed reference mode and
`r sec_open_fit_split_vs_db`% faster in open reference mode.
Thus, reference clustering with the split dataset strategy creates as high
quality OTUs as _de novo_ clustering yet at a faster run time, and fits far more
query sequences than the database strategy.

@@

```{r assert_split_fracs, include = FALSE}
close_enough(cv_fit_split_mcc_human_simple, 0, tol = 10^-2)
assert_that(sec_fit_split_human_simple_1 > sec_fit_split_human_simple_9)
assert_that(frac_fit_split_human_simple_1 < frac_fit_split_human_simple_9)
```

While we initially tested this strategy using a 50% split of the data into
reference and query fractions, we next investigated whether there was
an optimal reference fraction size.
To identify the best reference size, reference sets with 10% to 90% of the
sequences were created, with the remaining sequences used for the query
(Figure \ref{fig:results_split}).
OTU quality was remarkably consistent across reference fraction sizes.
For example, splitting the human dataset 100 times yielded a coefficient of
variation (i.e. the standard deviation divided by the mean) of
`r cv_fit_split_mcc_human_simple` for the MCC score across all fractions.
Run time generally decreased as the reference fraction increased; for the human
dataset, the median run time was `r sec_fit_split_human_simple_1` seconds with
10% of sequences in the reference and `r sec_fit_split_human_simple_9` seconds
with 90% of sequences in the reference.
The RAM usage was virtually the same across reference fraction sizes, with a
coefficient of variation of `r cv_fit_split_mem_human_simple` for the human
dataset (data not shown).
In closed reference mode, the fraction of sequences that mapped increased as the
reference size increased; for the human dataset, the median fraction mapped was
`r frac_fit_split_human_simple_1` with 10% of sequences in the reference and
`r frac_fit_split_human_simple_9` with 90% of sequences in the reference.
These trends held for the other datasets as well.
Thus, the reference fraction did not affect OTU quality in terms of MCC score
nor the memory usage, but did affect the run time and the fraction of sequences
that mapped during the closed reference clustering.
